---
title: "Fueling the Future: A Data-Driven Look at Historical Gasoline Demand in Ontario"
author: "Dylan Bui"
output: pdf_document
date: "2025-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

This project investigates the forecasting of monthly gasoline demand in Ontario, Canada, using historical data from January 1960 to December 1975. The primary objective is to construct a predictive model using the first 186 months of data (January 1960 - June 1975), and evaluate its forecasting performance over the final 6 months (July - December 1975).

To achieve this, a Box-Cox log transformation and differencing at lags 1 and 12 were applied to stabilize variance, and remove the trend and seasonality.  Various SARIMA models were then fitted, and the best model was selected by filtering through each of their AICc values, followed by diagnostic checking to confirm residual normality and white noise behavior. The final chosen model, SARIMA(2, 1, 0)(0, 1, 1)[12], produced forecast intervals that successfully captured the true observed values, with the exception of one outlier.  The forecasting results illustrate the model's effectiveness in capturing both seasonal and auto-regressive structure in gasoline demand, supporting SARIMA as a reliable method for short-term forecasting.


# Introduction

The dataset used in this project is titled "Monthly Gasoline Demand Ontario (gallon millions)", and was extracted from R’s Time Series Data Library (tsdl).  The data is recorded monthly, spanning from January 1960 to December 1975, and measures the amount of gasoline consumed or sold in Ontario, Canada, in millions of gallons.  My motivation for selecting this dataset was to observe how consumers and industries responded to economic cycles and price shocks, while also analyzing the seasonal patterns commonly associated with gasoline demand - patterns that may generalize to other regions.

The goals of this project are to capture any long-term trends and fluctuations between January 1960 and June 1975, fit a time series forecasting model using this training period, and generate a short-term forecast for the final six months, from July to December 1975.  To prepare the data, a Box-Cox log transformation was applied to achieve a stationary variance, followed by differencing at lags 1 and 12 to remove the trend and seasonal components.  A suitable ARMA model was generated by using maximum likelihood estimation, which implemented a for loop and compared models based on their AICc, however the model was too complex to use in analysis. Numerous SARIMA models, derived from the ACF and PACF of the transformed data, were then considered and evaluated.  The best model, model 5, was selected based on its low AICc value, stationarity, invertibility, and its performance on statistical diagnostic tests, including the Shapiro-Wilk, Box-Pierce, Ljung-Box, and McLeod-Li tests - all of which confirmed that the residuals resemble white noise.

While the AR(0) condition was not met, suggesting minor autocorrelation, this was acknowledged and investigated through multiple model refinements, which did not meaningfully reduce the autocorrelation.  Furthermore, the residual ACF and PACF plots consistently showed spikes at lag 23, which persisted despite adjustments to the seasonal structure.  Since these spikes did not appear to be seasonally related or significantly affect the residual diagnostics, they were concluded to be non-problematic, and model 5 was retained for forecasting.

The final selected model, SARIMA(2, 1, 0)(0, 1, 1)[12], generated forecasts whose confidence intervals captured five of the six actual observed values in the test set, with one point falling slightly outside.  While not all observed values were contained within the confidence bounds, the model still demonstrated reasonable and consistent forecasting performance.

All analysis was conducted in R, utilizing packages such as forecast, tsdl and stats.  The dataset was obtained from the Time Series Data Library, a publicly available resource in R.


# Import Data and Split Original Data into Train-Test Sets

```{r}

library(tsdl)

gas <- subset(tsdl, 12, "Sales")[[1]]
attr(gas, "description")

train_gas= gas[c(1:186)]
test_gas  = gas[c(187:192)]

train_gas
test_gas

```

# Plot Time Series of Training Data

```{r}

plot.ts(as.numeric(train_gas),
        main = "Original Time Series of Training Data",
        ylab = "Gas Demand")
ntr = length(as.numeric(train_gas))
fit_train <- lm(as.numeric(train_gas) ~ as.numeric(1:ntr))
abline(fit_train, col = "red")
abline(h = mean(as.numeric(train_gas)), col = "blue")

```

From the original time series plot, it is evident that the variance of the data increases as time increases, so we will first transform the data and stabilize the variance by applying a Box-Cox log transformation.


# Stabilize Variance with Box-Cox Log Transformation

```{r}

gas.log = log(train_gas)

par(mfrow = c(1,2))
hist(train_gas,
     col = "purple",
     xlab = "",
     main =" Histogram of Training Data")
hist(gas.log,
     col = "blue",
     xlab = "",
     main = "Histogram of Log(Gas) Data",
     breaks = 10)

```

By comparing the histogram before and after the log Box-Cox log transformation, the histogram of the log-transformed data appears more Gaussian.  Therefore, we will use the log-transformed data in the following analysis.

```{r}

gas.log_raw = as.numeric(gas.log)
plot.ts(gas.log_raw,
        main = "Log(Gas) Demand",
        ylab= "Log(Gas)")
nt = length(gas.log_raw)
fit <- lm(gas.log_raw ~ as.numeric(1:nt))
abline(fit, col = "red")
abline(h = mean(gas.log_raw), col = "blue")

```

The variance appears more stable as time passed, but the positive trend and seasonality is still clearly present in the data.  Before assessing whether or not applying the log-transformation to the time series is effective, I am going to remove the trend and seasonal components.


# ACF and PACF of Log-Transformed Data

```{r}

acf(gas.log, lag.max = 40, main = "")
title("ACF of Log(Gas)")
pacf(gas.log, lag.max = 40, main = "")
title("PACF of Log(Gas)")

```

There appears to be both trend and seasonality still present, with the ACF decaying slowly with time, while also having a period every 12 lags.  We will first deal with the trend, which can be removed by differencing at lag 1.


# Removing Trend Component of Log-Transformed Data

```{r}

gas.lag_1 <- diff(gas.log, lag = 1)
plot.ts(gas.lag_1,
        main = "Log(Gas), Differenced at Lag 1",
        ylab = "Log(Gas)")
fit_1 <- lm(gas.lag_1 ~ as.numeric(1:length(gas.lag_1)))
abline(fit_1, col = "red")
abline(h = mean(gas.lag_1), col = "blue")

```

After taking difference at lag 1 on the log-transformed data, we plot the differenced data with time. I can see that the trend disappears with mean almost equal to zero, with the red regression fit line (trend) appearing to be closely lined up to the blue horizontal line (constant mean). Thus, it can be concluded that differencing at lag 1 eliminates the trend, however, seasonality is still present.


# Comparing Variance After Removing Trend Component

```{r}

var_log <- var(gas.log)
var_diff_1 <- var(gas.lag_1)

```

The variance of the data also decreases from `r var_log` to `r var_diff_1`.  Therefore, differencing the log-transformed data at lag 1 was appropriate and effective in both removing the trend, and reducing the variance of the original data.


# ACF of Log-Transformed Data, Differenced at Lag 1

```{r}

acf(gas.lag_1, lag.max = 40, main = "")
title("ACF of Log(Gas), Differenced at Lag 1")

```

With a lower variance from differencing at lag 1, removing the trend component was necessary, however, the ACF still illustrates seasonality, which needs to be dealt with before a
fitted model can be chosen for forecasting.


# Removing Seasonal Component of Log-Transformed Data, Differenced at Lag 1

```{r}

gas.lag_1_12 <- diff(gas.lag_1, lag = 12)
plot.ts(gas.lag_1_12,
        main = "Log(Gas), Differenced at Lag 1 and 12",
        ylab = "Log(Gas)")
fit_1_12 <- lm(gas.lag_1_12 ~ as.numeric(1:length(gas.lag_1_12)))
abline(fit_1_12, col = "red")
abline(h = mean(gas.lag_1_12), col = "blue")
    
```

After taking the difference at both lag 1 and 12 on the log-transformed data, we plot the data with time.  It can be observed that the seasonality has been eliminated with no periods, with a stable variance over time that displays some peaks.  While the peaks stand out, this is likely attributed to random noise from real-world events that is non-problematic, so this data is sufficient for finding a potential fitted model.  Thus, we conclude that differencing at lag 12 eliminates the seasonality.


# Comparing Variance After Removing Seasonal Component

```{r}

var_diff_1 <- var(gas.lag_1)
var_diff_12 <- var(gas.lag_1_12)

```


After differencing at lag 12, the variance of the data decreases from `r var_diff_1` to `r var_diff_12`.  Therefore, differencing at lag 12 was also appropriate in helping to remove the seasonality, while again significantly reducing the variance.


# ACF and PACF of Log-Transformed Data, Differenced at Lag 1 and 12

```{r}

acf(gas.lag_1_12, lag.max = 60, main = "")
title("ACF of Log(Gas), Differenced at Lag 1 and 12")
pacf(gas.lag_1_12, lag.max = 60, main = "")
title("PACF of Log(Gas), Differenced at Lag 1 and 12")

```


# Selecting Coefficients For Potential Fitted Model

(p):
- While the ACF slowly decays, the PACF shows significant negative spike at lag 1 and 2, so I will consider both values.
- Thus, p = 1 or 2


(q):
- The ACF gradually decreases after lag 1, suggesting no MA model, however I will compare both values to ensure I fully capture the autocorrelation in the data.
- Thus, q = 0 or 1


(P):
- The PACF is not significant at lag 12 or 24, suggesting no seasonal AR term.
- Thus, P = 0


(Q):
- The ACF has a small significant negative spike at lag 24 and lag 36, which would equate to Q = 3.  However since less coefficients are more desirable, I am also going to consider Q = 2.
- Thus, Q = 2 or 3


# Find Suitable ARMA Model Using Maximum Likelihood Estimation


```{r, warning = FALSE}

AICc <- function(model) {
  loglik <- as.numeric(logLik(model))
  k <- attr(logLik(model), "df")
  n <- length(residuals(model))
  aic <- -2 * loglik + 2 * k
  aicc <- aic + (2 * k * (k + 1)) / (n - k - 1)
  return(aicc)
}

aiccs <- matrix(NA, nr = 6, nc = 6)
dimnames(aiccs) = list(p = 0:5, q = 0:5)
for(p in 0:5) {
for(q in 0:5) {
aiccs[p + 1, q + 1] = AICc(arima(gas.log, order = c(p, 1, q), method = "ML")) }
}

aiccs
(aiccs == min(aiccs))

```

Since I could not access the AICc() function in the qpcR library despite numerous efforts, I defined and created a function that performs identical calculations.  I used the function for maximum likelihood estimation, however my best model was ARMA(5, 5), which is not preferable since it has ten coefficients, a significantly large number.  Because of the complexity of this model, I will NOT be including it in my list of potential candidates for the best fitted model.


# Potential Models

```{r}

model1 <- arima(gas.log, order = c(1, 1, 0), seasonal = list(order = c(0, 1, 3), period = 12),
                method = "ML")
model2 <- arima(gas.log, order = c(1, 1, 0), seasonal = list(order = c(0, 1, 2), period = 12),
                method = "ML")
model3 <- arima(gas.log, order = c(1, 1, 1), seasonal = list(order = c(0, 1, 2), period = 12),
                method = "ML")
model4 <- arima(gas.log, order = c(2, 1, 1), seasonal = list(order = c(0, 1, 2), period = 12),
                method = "ML")

```


# Model Fitting

```{r}

# Model 1

model1 <- arima(gas.log, order = c(1, 1, 0), seasonal = list(order = c(0, 1, 3), period = 12), 
                method = "ML")
model1

```
Confidence Intervals

ar1: [-0.5636 - (2 * 0.0628), -0.5636 + (2 * 0.0628)]
= [-0.5636 - 0.1256, -0.5636 + 0.1256]
= [-0.6892, -0.438]

sma1: [-0.5398 - (2 * 0.1575), -0.5398 + (2 * 0.1575)]
= [-0.5398 - 0.315, -0.5398 + 0.315]
= [-0.8548, -0.2248]

sma2: [-0.4099 - (2 * 0.1068), -0.4099 + (2 * 0.1068)]
= [-0.4099 - 0.2136, -0.4099 + 0.2136]
= [-0.6235, -0.1963]

sma3: [-0.0501 - (2 * 0.1048), -0.0501 + (2 * 0.1048)]
= [-0.0501 - 0.2096, -0.0501 + 0.2096]
= [-0.2597, 0.1595]


From the calculated confidence intervals above, zero is within the interval for sma3, so it is NOT significant and I will drop it from model 1.


```{r}

# Model 2

model2 <- arima(gas.log, order = c(1, 1, 0), seasonal = list(order = c(0, 1, 2), period = 12), 
                method = "ML")
model2

```

Confidence Intervals

ar1: [-0.5620 - (2 * 0.0628), -0.5620 + (2 * 0.0628)]
= [-0.5620 - 0.1256, -0.5620 + 0.1256]
= [-0.6876, -0.4364]

sma1: [-0.5659 - (2 * 0.1999), -0.5659 + (2 * 0.1999)]
= [-0.5659 - 0.3998, -0.5659 + 0.3998]
= [-0.9657, -0.1661]

sma2: [-0.4339 - (2 * 0.1061), -0.4339 + (2 * 0.1061)]
= [-0.4339 - 0.2122, -0.4339 + 0.2122]
= [-0.6461, -0.2217]

From the calculated confidence intervals above, zero is NOT within the interval for all three coefficients.  Thus, all are significant and we do not drop any coefficients from model 2.

```{r}

# Model 3

model3 <- arima(gas.log, order = c(1, 1, 1), seasonal = list(order = c(0, 1, 2), period = 12), 
                method = "ML")
model3

```
Confidence Intervals

ar1: [-0.2642 - (2 * 0.0965), -0.2642 + (2 * 0.0965)]
= [-0.2642 - 0.193, -0.2642 + 0.193]
= [-0.4572, -0.0712]

ma1: [-0.7848 - (2 * 0.0725), -0.7848 + (2 * 0.0725)]
= [-0.7848 - 0.145, -0.7848 + 0.145]
= [-0.9298, -0.6398]

sma1: [-0.4563 - (2 *  0.0777), -0.4563 + (2 *  0.0777)]
= [-0.4563 - 0.1554, -0.4563 + 0.1554]
= [-0.6117, -0.3009]

sma2: [-0.2048 - (2 * 0.0747), -0.2048 + (2 * 0.0747)]
= [-0.2048 - 0.1494, -0.2048 + 0.1494]
= [-0.3542, -0.0554]

From the calculated confidence intervals above, zero is NOT within the intervals for all four coefficients.  Thus, they are all significant and we do not drop any coefficients from model 3.


```{r}

# Model 4

model4 <- arima(gas.log, order = c(2, 1, 1), seasonal = list(order = c(0, 1, 2), period = 12), 
                method = "ML")
model4

```
Confidence Intervals

ar1: [-0.9565 - (2 * 0.0868), -0.9565 + (2 * 0.0868)]
= [-0.9565 - 0.1736, -0.9565 + 0.1736]
= [-1.1301, -0.7829]

ar2: [-0.6589 - (2 * 0.0691), -0.6589 + (2 * 0.0691)]
= [-0.6589 - 0.1382, -0.6589 + 0.1382]
= [-0.7971, -0.5207]

ma1: [-0.0193 - (2 * 0.1310), -0.0193 + (2 * 0.1310)]
= [-0.0193 - 0.262, -0.0193 + 0.262]
= [-0.2813, 0.2427]

sma1: [-0.5289 - (2 * 0.08), -0.5289 + (2 * 0.08)]
= [-0.5289 - 0.16, -0.5289 + 0.16]
= [-0.6889, -0.3689]

sma2: [-0.0916 - (2 * 0.0822), -0.0916 + (2 * 0.0822)]
= [-0.0916 - 0.1644, -0.0916 + 0.1644]
= [-0.256, 0.0728]

From the calculated confidence intervals above, ma1 and sma2 are NOT significant since they include zero in their confidence intervals.  Thus, we will drop them from model 4, and apply this change in the next model.


```{r}

model5 <- arima(gas.log, order = c(2, 1, 0), seasonal = list(order = c(0, 1, 1), period = 12), 
                method = "ML")
model5

```
Confidence Intervals

ar1: [-0.9827 - (2 * 0.0553), -0.9827 + (2 * 0.0553)]
= [-0.9827 - 0.1106, -0.9827 + 0.1106]
= [-1.0933, -0.8721]

ar2: [-0.6875 - (2 * 0.0549), -0.6875  + (2 * 0.0549)]
= [-0.6875 - 0.1098, -0.6875 + 0.1098]
= [-0.7973, -0.5777]

sma1: [-0.5870 - (2 * 0.0726), -0.5870 + (2 * 0.0726)]
= [-0.5870 - 0.1452, -0.5870 + 0.1452]
= [-0.7322, -0.4418]


From the calculated confidence intervals above, zero is NOT within any of the intervals for all three coefficients.  Thus, all are significant and we do not drop any coefficients from model 5.


# AICc Comparison

```{r}

AICc <- function(model) {
  loglik <- as.numeric(logLik(model))
  k <- attr(logLik(model), "df")
  n <- length(residuals(model))
  aic <- -2 * loglik + 2 * k
  aicc <- aic + (2 * k * (k + 1)) / (n - k - 1)
  return(aicc)
}

AICc1 <- AICc(model1)
AICc1
AICc2 <- AICc(model2)
AICc2
AICc3 <- AICc(model3)
AICc3
AICc4 <- AICc(model4)
AICc4
AICc5 <- AICc(model5)
AICc5

```
I applied the AICc() function to each model to generate their respective AICc value, where model 5 had the lowest AICc value with the least amount of coefficients, making it the most ideal candidate for the best fitted model.  I will first check the stationarity, invertibility and diagnostic tests for model 5, and if they are insufficient, I will analyze additional models.

# Check if Model 5 is Stationary and Invertible

```{r}

model5 <- arima(gas.log, order = c(2, 1, 0), seasonal = list(order = c(0, 1, 1), period = 12), 
                method = "ML")
model5

```

Formula for Model 5:
(1 + 0.9827B + 0.6875B^2)(1 - B)(1 - B^12)Xt = (1 - 0.5870B^12)Zt,
and Xt = log(Ut), where Ut was the original data


- SAR part: No SAR part

- AR part: The equation 1 + 0.9827z + 0.6875z^2 = 0 has roots -0.7146909+0.9714743i and -0.7146909-0.9714743i, which converts to 1.2060.  Since |1.2060| > 1, the model 5 is stationary.

```{r}

polyroot(c(1, 0.9827, 0.6875))

```

- SMA part: Since |-0.5870| < 1, model is also invertible.

- MA part: No MA part

Therefore, model 5 is indeed both stationary and invertible.


# Diagnostic Checking of Model 5

```{r}

res5 = residuals(model5)
par(mfrow = c(2,2))
hist(res5, density = 20, breaks = 20, col = "blue", xlab = "", prob = TRUE,
     main = "Histogram of Residuals of Model 5")
m5 <- mean(res5)
std5 <- sqrt(var(res5))
curve(dnorm(x, m5, std5), add = TRUE )
plot.ts(res5, ylab = "Residuals of Model 5", main = "Residuals Plot of Model 5")
fitt5 <- lm(res5 ~ as.numeric(1:length(res5)))
abline(fitt5, col = "red")
abline(h = mean(res5), col = "blue")
qqnorm(res5, main = "Normal Q-Q Plot for Model 5")
qqline(res5, col="blue")

```

Based on the histogram and normal Q-Q plot for model 5, the residuals do appear normal, but we will conduct more accurate testing to ensure this assumption is true.  In the time series plot of the residuals, again there is no trend or obvious seasonality.  Now, we will check the Portmanteau Statistics, where we still want the p-values for all tests to be above 0.05.

# Portmanteau Statistics for Model 5

```{r}

# Shapiro test for Normality
shapiro.test(res5)

# Box-Pierce Test
Box.test(res5, type = c("Box-Pierce"), lag = 13, fitdf = 3)

# Ljung-Box test
Box.test(res5, type = c("Ljung-Box"), lag = 13, fitdf = 3)

# McLeod-Li test
Box.test((res5)^2, type = c("Ljung-Box"), lag = 13, fitdf = 0)

```

lag: Here, we choose lag = 13 because we have n = 186 observations and the square root is about 13.
fitdf: Also, model 5 has 3 estimated coefficients, so we set fitdf = 3.

We can see that the p-values are larger than 0.05 for all the Portmanteau Statistics tests, so model 5 does pass all tests.  Next, I will check if ar() function recommends order 0 result for the residuals, which means that the residuals are white noise for the PACF.

# Check if the residuals are AR(0) for Model 5

```{r}

ar(res5, aic = TRUE, order.max = NULL, method = c("yule-walker"))

```
The residuals do NOT select order 0 for the residuals, which is indicative that the residuals are not white noise and there is some minor autocorrelation still present.  I will still proceed with plotting the ACF and PACF of the residuals.


# ACF and PACF of Residuals for Model 5

```{r}

acf(res5, lag.max = 40, main = "")
title("ACF of the Residuals of Model 5")
pacf(res5, lag.max = 40, main = "")
title("PACF of the Residuals of Model 5")

```
Looking at both the residual ACF and PACF for my best model, it appears that a seasonal term might need to be added from the correlation present at lag 23.  However, I considered this by altering the seasonal terms for P and Q, and checking their respective residual ACF and PACF, where the spikes at lag 23 were never eliminated.  The rest of the spikes in both the ACF and PACF can be considered borderline cases and are isolated, so they are likely random noise and are non-problematic.  Since adding seasonal values had no effect on the diagnostic checking analysis, and a simpler model is more desirable for forecasting, I decided that model 5, or SARIMA(2, 1, 0)(0, 1, 1)[12] is the best fitted model. While the model did pass all the Portmanteau Statistics tests for the ACF, it did not meet the AR(0) condition in regards to the PACF.  The model was adjusted multiple times to account for this autocorrelation still present, however, no model was successful in eliminating it completely, so this model has been chosen as the best fitted model.  The model was also stationary, invertible, and had the lowest AICc, so despite not fulfilling all tests, I will proceed with forecasting.


# Best Model For Forecasting

```{r}

bestmodel <- arima(gas.log,
                   order = c(2, 1, 0),
                   seasonal = list(order = c(0, 1, 1),
                   period = 12),
                   method = "ML")
bestmodel

```

# Forecasting 

In the forecasting section, we will now predict the gasoline demand from July to December 1975 based on our model, and compare these prediction points with the true values.

```{r, message = FALSE}

library(forecast)

pred.log <- predict(bestmodel, n.ahead = 6)

# Compute confidence intervals on log scale
U.tr = pred.log$pred + 2 * pred.log$se
L.tr = pred.log$pred - 2 * pred.log$se

# Plot of predicted points on log scale
ts.plot(gas.log, xlim = c(1, length(gas.log) + 6), ylim = c(min(gas.log), max(U.tr)))
lines(U.tr, col = "blue", lty = "dashed")
lines(L.tr, col = "blue", lty = "dashed")
points((length(gas.log) + 1):(length(gas.log) + 6), pred.log$pred, col = "black", pch = 16)

# Zoomed-in plot on log scale
ts.plot(gas.log, xlim = c(150, length(gas.log) + 6), ylim = c(12, max(U.tr)))
lines(U.tr, col = "blue", lty = "dashed")
lines(L.tr, col = "blue", lty = "dashed")
points((length(gas.log) + 1):(length(gas.log) + 6), pred.log$pred, col = "black", pch = 16)


```
 
In the forecast plot on the log-transformed data, the six-step ahead forecasted values are within the calculated confidence intervals.  Now, I will convert the data back to the original scale before applying the Box-Cox log transformation, and then fit my best fitted model on the test set, which are the last six months spanning from July to December 1975.

# Fit the Best Model on the Test Set

```{r}

pred.log <- predict(bestmodel, n.ahead = 6)

m.log <- pred.log$pred
se.log <- pred.log$se

# Compute confidence intervals
U.log = m.log + (2 * se.log)
L.log = m.log - (2 * se.log)

# Convert confidence intervals to original scale
pred.orig <- exp(m.log)
U <- exp(U.log)
L <- exp(L.log)

index <- (length(train_gas) + 1) : (length(train_gas) + length(test_gas))

# Plot on original scale with test set
ts.plot(as.numeric(gas), ylim = c(0, max(U)), col = "red",
ylab = "Gas Demand", main = "Visualization of Forecasting on Test Set")
lines(index, U, col = "blue", lty = "dashed")
lines(index, L, col = "blue", lty = "dashed")
points(index, pred.orig, col = "black", pch = 16)
points(index, test_gas, col = "purple", pch = 16)

# Zoomed-in plot on original scale with test set
ts.plot(as.numeric(gas), xlim = c(180, length(train_gas) + 6), ylim = c(200, max
(U)), col = "red", ylab = "Gas Demand", main = "Zoomed-In Visualization of Forecasting on Test Set")
lines(index, U, col = "blue", lty = "dashed")
lines(index, L, col = "blue", lty = "dashed")
points(index, pred.orig, col = "black", pch = 16)
points(index, test_gas, col = "purple", pch = 16)

```

From the zoomed-in plot of the forecast on the test set, it is evident that one of the actual observed values (shown in purple) falls slightly outside the calculated prediction interval.  While this indicates that the model may not have fully captured the underlying structure of the data, such deviations can occur for various reasons, including real-world events not reflected in historical patterns. One possible explanation could be the 1973 Oil Crisis, which led to an exponential increase in global oil prices and may have had a substantial effect on gasoline consumption demand in Ontario during the forecast period.  Although the crisis began in 1973, its economic consequences that included fuel rationing and price controls most likely persisted into subsequent years.  Thus, this minor forecasting inaccuracy could potentially be attributed to a real-world event that disrupted typical seasonal or trend behavior, rather than model misspecification alone. The observed deviation could have also been caused by other factors, such as local weather anomalies or random forecast error.  However, given the historical volatility of energy markets, a single point falling outside the prediction interval is expected.


# Conclusion

The primary goals of this project were to analyze long-term patterns in gasoline demand in Ontario between Januarsy 1960 and June 1975, fit a reliable time series model for forecasting, and evaluate its predictive accuracy over the final six months, from July to December 1975.  These goals were sufficiently achieved. The final selected fitted model, SARIMA(2, 1, 0)(0, 1, 1)[12], performed well in capturing the underlying trend and seasonal components of the data, producing prediction intervals that ultimately contained five out of the six actual observed values in the test set.

Despite a minor residual autocorrelation issue and persistent spikes at lag 23 in the residual ACF and PACF, the best fitted model passed all major diagnostic checks for the ACF, including the Shapiro-Wilk, Box-Pierce, Ljung-Box, and McLeod-Li tests.  The issues related to the AR(0) condition and lag 23 behavior were further investigated through alternate models, but due to a lack of identifiable seasonal structure, they were considered non-problematic.

The mathematical form of the final model is:

(1 + 0.9827B + 0.6875B^2)(1 - B)(1 - B^12)Xt = (1 + 0.5870B^12)Zt, 

where Xt = log(Ut), B is the backshift operator and Zt represents white noise.

All computations and analyses were carried out in R, using packages such as forecast, tsdl, and functions from the stats package.

I would like to thank Professor Feldman for her guidance throughout this course and project, as well as my fellow classmate for helping me find this time series dataset to analyze.  I also appreciate the support of teaching assistant, Lihao Xiao, who clarified several aspects of the residual diagnostic analysis during office hours.


# References

R Core Team. (2024). R: A Language and Environment for Statistical Computing (Version 2024.09.0+375) [Computer software]. R Foundation for Statistical Computing.
Koivu, J. (2024). tsdl: Time Series Data Library (Version 0.1.0) [R package].
Hyndman, R.J., and Khandakar, Y. (2024). forecast: Forecasting functions for time series and linear models (Version 8.22.0) [R package].


# Appendix

```{r, echo = TRUE, eval = FALSE}

# Load required packages
library(tsdl)      # Access the Time Series Data Library
library(forecast)  # For predict()


# Import the dataset "Monthly gasoline demand Ontario gallon millions 1960 – 1975"
# 12 is the domain
# 'Sales' is the category
# [[1]] represents position in the list
gas <- subset(tsdl, 12, "Sales")[[1]]


# Print description of the data set
attr(gas, "description")


# Split data into training (first 186 months) and test (last 6 months)
train_gas <- gas[1:186] # January 1960 to June 1975
test_gas  <- gas[187:192] # July 1975 to December 1975


# Plot training data, with trend and mean lines
plot.ts(as.numeric(train_gas), # Plot original data
        main = "Original Time Series of Training Data",
        ylab = "Gas Demand")
ntr = length(as.numeric(train_gas))
fit_train <- lm(as.numeric(train_gas) ~ as.numeric(1:ntr))
abline(fit_train, col = "red") # Plot trend of original data
abline(h = mean(as.numeric(train_gas)), col = "blue") # Plot mean of original data over time


# Stabilize variance by Box-Cox log transformation
gas.log = log(train_gas) # Apply log transformation to original data
par(mfrow = c(1,2))
hist(train_gas, # Plot histogram of original data
     col = "purple",
     xlab = "",
     main = " Histogram of Training Data")
hist(gas.log, # Plot histogram of log-transformed data
     col = "blue",
     xlab = "",
     main = "Histogram of Log-Transformed Data", breaks = 10)


# Plot log‐transformed data with trend and mean lines
gas.log_raw = as.numeric(gas.log)
plot.ts(gas.log_raw, # Plot log-transformed data
        main = "Log(Gas) Demand",
        ylab = "Log(Gas)")
nt = length(gas.log_raw)
fit <- lm(gas.log_raw ~ as.numeric(1:nt))
abline(fit, col = "red") # Plot trend of log-transformed data
abline(h = mean(gas.log_raw), col = "blue") # Plot mean of log-transformed data over time


# Plot ACF and PACF of log‐transformed data
acf(gas.log, lag.max = 40, main = "")
title("ACF of Log(Gas)")
pacf(gas.log, lag.max = 40, main = "")
title("PACF of Log(Gas)")


# Remove trend component by differencing log-transformed data at lag 1
gas.lag_1 <- diff(gas.log, lag = 1) # Difference at lag 1, which removes the trend component
plot.ts(gas.lag_1,  # Plot the data
        main = "Log(Gas), Differenced at Lag 1",
        ylab = "Log(Gas)")
fit_1 <- lm(gas.lag_1 ~ as.numeric(1:length(gas.lag_1)))
abline(fit_1, col = "red") # Plot trend of lag 1 differenced data
abline(h = mean(gas.lag_1), col = "blue") # Plot mean of lag 1 differenced data over time


# Compare variance before and after differencing at lag 1
var_log <- var(gas.log) # Variance of log‐transformed data
var_diff_1 <- var(gas.lag_1) # Variance after differencing at lag 1


# Plot ACF of log-transformed data, differenced at lag 1
acf(gas.lag_1, lag.max = 40, main = "")  
title("ACF of Log(Gas), Differenced at Lag 1")


# Remove seasonal component by also differencing at lag 12
gas.lag_1_12 <- diff(gas.lag_1, lag = 12) # Difference at lag 12, which removes the seasonal component
plot.ts(gas.lag_1_12,  # Plot the data
        main = "Log(Gas),
        Differenced at Lag 1 and 12",
        ylab = "Log(Gas)")
fit_1_12 <- lm(gas.lag_1_12 ~ as.numeric(1:length(gas.lag_1_12)))
abline(fit_1_12, col = "red")
abline(h = mean(gas.lag_1_12), col = "blue")

# Compare variance before and after differencing at lag 12
var_diff_1 <- var(gas.lag_1)
var_diff_12 <- var(gas.lag_1_12) # Confirm smaller variance


# Plot ACF and PACF of log-transformed data, differenced at lag 1 and 12
acf(gas.lag_1_12, lag.max = 40, main = "") # Plot ACF of data, differenced at lag 1 and 12
title("ACF of Log(Gas), Differenced at Lag 1 and 12")
pacf(gas.lag_1_12, lag.max = 40, main = "") # Plot PACF of data, differenced at lag 1 and 12
title("PACF of Log(Gas), Differenced at Lag 1 and 12")


# Define and create a AICc() function since could not access AICc() function from library(qpcR)
# Performs identical calculations
# Uses AICc formula: AICc = AIC + (2k * (k + 1)) / (n - k - 1),
# where AIC =  -2 * log-likelihood + 2k,
# k is number of estimated parameters
# and n is number of observations to fit
AICc <- function(model) {
  loglik <- as.numeric(logLik(model)) # logLik() function obtains log-likelihood
  k <- attr(logLik(model), "df") # returns number of parameters
  n <- length(residuals(model)) # returns number of observations used to fit
  aic <- -2 * loglik + 2 * k
  aicc <- aic + (2 * k * (k + 1)) / (n - k - 1)
  return(aicc)
}


# Use maximum likelihood estimation to find suitable ARMA model
# Generated ARMA(5, 5), which was not considered due to complexity
aiccs <- matrix(NA, nr = 6, nc = 6)
dimnames(aiccs) = list(p = 0:5, q = 0:5)
for(p in 0:5) {
for(q in 0:5) {
aiccs[p + 1, q + 1] = AICc(arima(gas.log, order = c(p, 1, q), method = "ML")) }
}
aiccs
(aiccs == min(aiccs))


# Fit potential models into the arima() function
# Print out each model and examine coefficient values to calculate confidence intervals
# Approximate 95% confidence intervals for each model were computed using the formula:
# Coefficient value ± (2 * SE)
model1 <- arima(gas.log, 
                order = c(1, 1, 0), 
                seasonal = list(order = c(0, 1, 3), period = 12), 
                method = "ML")
model1
model2 <- arima(gas.log, 
                order = c(1, 1, 0), 
                seasonal = list(order = c(0, 1, 2), period = 12), 
                method = "ML")
model2
model3 <- arima(gas.log, 
                order = c(1, 1, 1), 
                seasonal = list(order = c(0, 1, 2), period = 12), 
                method = "ML")
model3
model4 <- arima(gas.log, 
                order = c(2, 1, 1), 
                seasonal = list(order = c(0, 1, 2), period = 12), 
                method = "ML")
model4
model5 <- arima(gas.log, 
                order = c(2, 1, 0), 
                seasonal = list(order = c(0, 1, 1), period = 12), 
                method = "ML")
model5  # Model 4, but without ma1 and sma2


# Compare AICc for each model
# Model 5 selected as best fitted model
# Proceed with testing to ensure this is true
AICc1 <- AICc(model1)
AICc1
AICc2 <- AICc(model2)
AICc2
AICc3 <- AICc(model3)
AICc3
AICc4 <- AICc(model4)
AICc4
AICc5 <- AICc(model5) 
AICc5 # Lowest AICc


# Check stationarity of model 5
# Invertibility required no computation
polyroot(c(1, 0.9827, 0.6875))


# Diagnostic plots for model 5
res5 = residuals(model5)
par(mfrow = c(2,2))
hist(res5, density = 20, breaks = 20, col = "blue", xlab = "", prob = TRUE,
     main = "Histogram of Residuals of Model 5")
m5 <- mean(res5)
std5 <- sqrt(var(res5))
curve(dnorm(x, m5, std5), add = TRUE )
plot.ts(res5, ylab = "Residuals of Model 5", main = "Residuals Plot of Model 5")
fitt5 <- lm(res5 ~ as.numeric(1:length(res5)))
abline(fitt5, col = "red")
abline(h = mean(res5), col = "blue")
qqnorm(res5, main = "Normal Q-Q Plot for Model 5")
qqline(res5, col="blue")


# Portmanteau statistic tests for model 5

# Shapiro test for Normality
shapiro.test(res5)
# Box-Pierce Test
Box.test(res5, type = c("Box-Pierce"), lag = 13, fitdf = 3)
# Ljung-Box test
Box.test(res5, type = c("Ljung-Box"), lag = 13, fitdf = 3)
# McLeod-Li test
Box.test((res5)^2, type = c("Ljung-Box"), lag = 13, fitdf = 0)


# Check if residuals meet AR(0) condition for model 5
ar(res5, aic = TRUE, order.max = NULL, method = c("yule-walker"))


# Plot ACF and PACF of model 5 residuals
acf(res5, lag.max = 40, main = "")
title("ACF of the Residuals of Model 5")
pacf(res5, lag.max = 40, main = "")
title("PACF of the Residuals of Best Model 5")


# Re-establish the best model (model 5)
bestmodel <- arima(gas.log,
                   order = c(2, 1, 0),
                   seasonal = list(order = c(0, 1, 1),
                   period = 12),
                   method = "ML")
bestmodel


# Forecast and plot six steps (months) ahead on log scale
library(forecast)
# Use predict() function to forecast the last six months
pred.log <- predict(bestmodel, n.ahead = 6)
# Compute confidence intervals on log scale
U.tr = pred.log$pred + 2 * pred.log$se
L.tr = pred.log$pred - 2 * pred.log$se
# Plot the predicted points on log scale
ts.plot(gas.log, xlim = c(1, length(gas.log) + 6), ylim = c(min(gas.log), max(U.tr)))
lines(U.tr, col = "blue", lty = "dashed")
lines(L.tr, col = "blue", lty = "dashed")
points((length(gas.log) + 1):(length(gas.log) + 6), pred.log$pred, col = "black", pch = 16)
# Zoomed-in plot of predict points on log scale
# Black dots = predicted values
ts.plot(gas.log, xlim = c(150, length(gas.log) + 6), ylim = c(12, max(U.tr)))
lines(U.tr, col = "blue", lty = "dashed")
lines(L.tr, col = "blue", lty = "dashed")
points((length(gas.log) + 1):(length(gas.log) + 6), pred.log$pred, col = "black", pch = 16)


# Fit the best model on the test set
pred.log <- predict(bestmodel, n.ahead = 6)
# Compute values for confidence intervals
m.log <- pred.log$pred
se.log <- pred.log$se
# Compute confidence intervals
U.log = m.log + (2 * se.log)
L.log = m.log - (2 * se.log)
# Convert back to original scale
pred.orig <- exp(m.log)
U <- exp(U.log)
L <- exp(L.log)
# Create index to include test set (actual values)
index <- (length(train_gas) + 1) : (length(train_gas) + length(test_gas))
# Plot on original scale and implement the test set
ts.plot(as.numeric(gas), ylim = c(0, max(U)), col = "red",
ylab = "Gas Demand", main = "Visualization of Forecasting on Test Set")
lines(index, U, col = "blue", lty = "dashed")
lines(index, L, col = "blue", lty = "dashed")
points(index, pred.orig, col = "black", pch = 16)
points(index, test_gas, col = "purple", pch = 16)
# Zoomed-in plot to compare true values to predicted values
# Black dots = predicted values
# Purple dots = true values
ts.plot(as.numeric(gas), xlim = c(180, length(train_gas) + 6), ylim = c(200, max
(U)), col = "red", ylab = "Gas Demand", main = "Zoomed-In Visualization of Forecasting on Test Set")
lines(index, U, col = "blue", lty = "dashed")
lines(index, L, col = "blue", lty = "dashed")
points(index, pred.orig, col = "black", pch = 16)
points(index, test_gas, col = "purple", pch = 16)

```

